The Precambrian paleogeography of Laurentia

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add ECMB pole to table and a reference to it in the text; add new paleogeo reconstruction figure

Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean.

On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 550 Ma \citep{Dickerson1998a, Martin2019a}. Rifting of the Laurentia margin north of the New York promontory was occuring by ca. 560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{}Hodych2007.

@@

Following the amalgamation of the Archean provinces in Laurentia by ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig8_Grenville_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic modified from \cite{Swanson-Hysell2019a}}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction envisions the Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018b}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

Magmatism ca. 615 Ma in both Laurentia and Baltica (in the relative configuration shown in Figure \ref{fig:Grenville_reconstructions}) is interpreted to be associated with the plume-related Central Iapetus Magmatic Province (CIMP; \citealp{Tegner2019a}). This ca. 615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Fig. \ref{fig:tectonic_history}; \citealp{Cawood2001a, Tegner2019a}). The cratons would once again be conjoined ca. 430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. In this model, North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} pointed to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma Narssaq Gabbro and Hvidaal dike poles during which it ascends to the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). In its reconstructed position, the ca. 1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwestern Laurentia margin prior to 1220 Ma. In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwestern Laurentia.

@@

>>>>

Extensive rifting followed by thermal subsidence occurred along the southeastern to eastern Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 540 Ma \citep{Dickerson1998a, Martin2019a}. Rifting of the Laurentia margin north of the New York promontory was occuring by ca. 560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{Hodych2007a}.

@@

Following the amalgamation of the Archean provinces in Laurentia by ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton \citep{Swanson-Hysell2021b}. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently been separated through translation and rifting. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging and non-unique given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland and those of East Svalbard \citep{Maloof2006a, Hoffman2012a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig8_Rodinia_Reconstruction.pdf}
\caption{\textbf{Paleogeographic reconstructions of Laurentia and other select conjugate Proterozoic continents leading up to Rodinia assembly in the late Mesoproterozoic and to its initial break-up in the Neoproterozoic}. The hypothesized connection between Siberia and Laurentia is implemented following \cite{Evans2016b} who interpret this relationship as persistent from 1.7 to 0.7 Ga. The reconstruction of North China to Laurentia following \cite{Ding2021a}. The reconstruction implements Kalahari and Amazonia cratons as conjugates with Laurentia in the Grenvillian orogeny \citep{Hoffman1991a}. The Australia-East Antarctica relationship with Laurentia follows \cite{Swanson-Hysell2012a} and is similar to the Neoproterozoic reconstruction between the continents of \cite{Li2011a} and implementing their relative rotation of North Australia relative to South and West Australia. This configuration back to ca. 1140 Ma is consistent with a comparison between the Laurentia poles of that age and the coeval Mt. Isa dikes pole from North Australia and with the Keweenawan Track if the the Nonesuch and Freda poles are interpreted to be ca. 1080 (consistent with chronostratigraphic constraints; \citealp{Slotznick2018b}) with further motion by ca. 1070 Ma. The time slices show the rapid motion of Laurentia implied by the paleomagnetic poles which is consistent with the timing of collisional orogenesis associated with the Grenvillian orogeny. The assembled Rodinia persisted until initial rifting ca. 775 Ma with episodic rifting continuing until ca. 530 Ma.}
\label{fig:Grenville_reconstructions}
\end{figure*}

@@

Magmatism ca. 615 Ma in both Laurentia and Baltica (in the relative configuration shown in Figure \ref{fig:Grenville_reconstructions}) is interpreted to be associated with the plume-related Central Iapetus Magmatic Province (CIMP; \citealp{Tegner2019a}). This ca. 615 to 590 Ma magmatism is hypothesized to have initiated the breakup between Baltica and Laurentia that led to the opening of the of the Iapetus Ocean (Figs. \ref{fig:tectonic_history} and \ref{fig:Grenville_reconstructions}; \citealp{Cawood2001a, Tegner2019a}). The cratons would once again be conjoined ca. 430 Ma during their collision that led to the Caledonian orogeny resulting in a continent that is referred to as Laurussia \citep{Torsvik2017a}.

@@

The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP \citep{Zhao2019a, Ding2021a, Zhang2021a}. While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} proposed that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwestern margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. In this model, North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. A challenge with interpreting these North China poles as primary is that they include data from limestones that would reconstruct them to have been deposited at very high latitude ($>$80\textdegree\ for the limestones of the lower member of the Nanfen Formation). Such high-latitude limestones would require non-uniformitarian depositional conditions for carbonate precipitation which is more dominant in low latitudes due to the temperature dependence of carbonate saturation. As additional support for a North China--Laurentia connection, \cite{Zhao2019a} pointed to similarities in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins as supporting this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival adjacent to Laurentia is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the 1237 $\pm$ 5 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and \cite{Zhang2021a} to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. Laurentia's pole path has a gap from the 1237 $\pm$ 5 Ma Sudbury pole to the 1184 $\pm$ 5 Ma Narssaq Gabbro and Hvidaal dike poles during which it ascends to the Logan Loop (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). In its reconstructed position, the ca. 1220 Ma North China pole is close to this gap such that it is possible that the APWPs are consistent and that North China arrived on the northwestern Laurentia margin prior to 1220 Ma. In terms of departing from this position, one possibility is that its departure is associated with latest Neoproterozoic extension in northwestern Laurentia.

@@

                 Laurentia &                                   East Central Minnesota Batholith &     1779$^{+2}_{-2}$ &      NR &     265.8 &      45.5 & 265.8 &  20.4 &      4.5 &        63.5 &                                   \cite{Swanson-Hysell2021b} \\ \hline
